---
title: "COVID-19 Ontario Schools Exploratory Data Analysis"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
# DEPENDENCIES -----------------------------------------------------------------

suppressPackageStartupMessages(expr = {
	
	library(plotly)
	library(magrittr)
	
	# load all data
	source('data_downloader.R')	
	
})

# SCRIPT SETTINGS --------------------------------------------------------------

data_length_days <- 180L

data_start_date <- min(covid19_schools_active$reported_date)

message('data starts on ', data_start_date)
data_end_date <- data_start_date + days(data_length_days)
message('data ends on ', data_end_date)

# LOAD DATA --------------------------------------------------------------------

## load policy tracing data set ------------------------------------------------
policy_tracing <- read.csv('data/policy_tracing.csv', sep = '|')
policy_tracing$Date <- as.Date(policy_tracing$Date, format =  '%d %b %Y')

# CLEAN DATA -------------------------------------------------------------------

```

# Source Data

## Student Demographic Data

See: <https://data.ontario.ca/dataset/d85f68c5-fcb0-4b4d-aec5-3047db47dcd5/resource/602a5186-67f5-4faf-94f3-7c61ffc4719a/download/new_sif_data_table_2018_2019prelim_en_august.xlsx>

```{r echo=FALSE, warning=FALSE, message=FALSE}
write.csv2(school_demographics, "./school_demographics.csv")

readLines("./school_demographics.csv") %>% 
	paste0(collapse="\n") %>% 
	openssl::base64_encode() -> encoded
```

[Download Student Demographics CSV](`r sprintf('data:text/csv;base64,%s', encoded)`)

## Combined Active Cases and Student Demographic Data

See: <https://data.ontario.ca/dataset/b1fef838-8784-4338-8ef9-ae7cfd405b41/resource/8b6d22e2-7065-4b0f-966f-02640be366f2/download/schoolsactivecovid.csv>

```{r echo=FALSE, warning=FALSE, message=FALSE}
write.csv2(cases_per_school, "./cases_per_school.csv")

readLines("./cases_per_school.csv") %>% 
	paste0(collapse="\n") %>% 
	openssl::base64_encode() -> encoded
```

[Download School Cases and Demographics CSV](`r sprintf('data:text/csv;base64,%s', encoded)`)

# Relationship Between Poverty Rates and Infection Rates

## Empirical Distribution of Poverty Rates

```{r echo=FALSE, warning=FALSE, message=FALSE}

# EXPLORATORY DATA ANALYSIS ----------------------------------------------------

## relationship between poverty and infection rates ----------------------------

### create poverty rates categorical variable and analyze ----------------------
poverty_rates <- as.integer(school_demographics$`percentage of school-aged children who live in low-income households`)
hist(poverty_rates, 
	 main = 'Province-Wide Empirical Distribution of Poverty Rates', 
	 xlab = 'poverty rate (%)')

summary(poverty_rates) # median = 15%

high_poverty_rate_threshold <- 15 # percentage of students living in poverty to define a "high poverty" school

```

We create the `poverty_rate_categorical` variable using a `r high_poverty_rate_threshold`% of children living in low-income households as a threshold for identification of high-poverty institutions in the active cases data set.

The proportion of low and high poverty institutions in the active cases data set is as follows:

```{r echo=FALSE, warning=FALSE, message=FALSE}
poverty_rate_categorical <- cut(cases_per_school$low_income, 
								breaks = c(0, high_poverty_rate_threshold, 100), 
								labels = c('low', 'high'))
prop.table(table(poverty_rate_categorical))

```

## Infection Rates

We define the `infection_rates` variable by dividing the number of active cases in a given school by the enrolment specified in the student demographics data set: $$ infection\_rate = cases / enrolment$$
This is a very crude measure as current school enrolments may vary widely from the values stated in the student demographics data set. One can operate on the assumption that the enrolment figures cited in the demographic data set may actually be inflated due to  school closures, remote learning, etc.

```{r echo=FALSE, warning=FALSE, message=FALSE}
### create infection rates categorical variable and analyze --------------------
infection_rates <- cases_per_school$cases_per_school / cases_per_school$school_enrolment
hist(infection_rates, main = 'empirical distribution of infection rates', xlab = 'infection rate')

summary(infection_rates)

high_infection_rate_threshold <- as.numeric(quantile(infection_rates, na.rm = TRUE, probs = 0.75))
```

We create the `infection_rate_categorical` variable using `r round(high_infection_rate_threshold, 4)` (which represents the 75th percentile of the empirical distribution of infection rates) as a threshold for identification of institutions with a high rate of infections in the active cases data set.

The proportion of low and high infection rate institutions in the active cases data set is as follows:

```{r echo=FALSE, warning=FALSE, message=FALSE}

infection_rate_categorical <- cut(infection_rates, 
								  breaks = c(0, high_infection_rate_threshold, 1), 
								  labels = c('low', 'high'))
prop.table(table(infection_rate_categorical))
```

Here are the case counts and proportions by `infection_rate_categorical`:

```{r echo=FALSE, warning=FALSE, message=FALSE}
### plot case counts vs poverty rates categorical ------------------------------
tbl <- tapply(cases_per_school$cases_per_school, 
			  poverty_rate_categorical,
			  sum)
barplot(tbl, main = 'Case Counts by Poverty Rate Categorical')
print(tbl)

### plot infection rates vs poverty rates categorical --------------------------
tbl <- tapply(infection_rates, 
			  poverty_rate_categorical,
			  median)
barplot(tbl, main = 'Median Infection Rates by Poverty Rate Categorical')
print(tbl)

```

## Relationship Between Poverty Rate and Infection Rate Categorical Variables

We now perform Chi-squared test to determine if the difference in rates is in fact significant. Here is our contingency table:

```{r echo=FALSE, warning=FALSE, message=FALSE}
# is the relationship between infection rates and poverty rates statistically significant?

### create contingency table ---------------------------------------------------
tbl_contingency <- table(infection_rate_categorical,
						 poverty_rate_categorical)
print(tbl_contingency)
```

Here are the results of the Chi-squared test:

```{r echo=FALSE, warning=FALSE, message=FALSE}

### perform chi-squared test ---------------------------------------------------
chisq_results <- chisq.test(tbl_contingency)
print(chisq_results)
# print(chisq_results$expected)
# print(chisq_results$observed)
```

We therefore reject the null hypothesis that the `infection_rate_categorical` and `poverty_rate_categorical` variables are independent of one another.

```{r echo=FALSE, warning=FALSE, message=FALSE}
# ## relationship between school enrolment and infection rates -------------------
# 
# ### analyze school enrolment ---------------------------------------------------
# # table(school_demographics$`school level`)
# idx <- which(school_demographics$`school level` == 'Elementary')
# enrolment <- as.integer(school_demographics$enrolment[ idx ])
# hist(enrolment, 
# 	 main = 'elementary school enrolment empirical distribution')
# summary(enrolment)
# enrolment_primary_cuts <- as.integer(quantile(enrolment, na.rm = TRUE))[ 2:5 ]
# idx <- which(school_demographics$`school level` == 'Secondary')
# enrolment <- as.integer(school_demographics$enrolment[ idx ])
# hist(enrolment, 
# 	 main = 'secondary school enrolment empirical distribution')
# summary(enrolment)
# enrolment_secondary_cuts <- as.integer(quantile(enrolment, na.rm = TRUE))[ 2:5 ]
# 
# ### cases by enrolment elementary ----------------------------------------------
# idx <- which(cases_per_school$school_level == 'Elementary')
# enrolment_elementary <- as.integer(cases_per_school$school_enrolment[ idx ])
# enrolment_elementary_categorical <- cut(enrolment_elementary, 
# 										breaks = enrolment_primary_cuts, 
# 										labels = c('small', 'medium', 'large'))
# case_count_elementary <- as.integer(cases_per_school$cases_per_school[ idx ])
# tapply(case_count_elementary, 
# 	   enrolment_elementary_categorical, 
# 	   sum) %>%
# 	barplot(main = 'infection counts by elementary school size')
# tapply(case_count_elementary / enrolment_elementary, 
# 	   enrolment_elementary_categorical, 
# 	   median) %>% 
# 	barplot(main = 'infection rates by elementary school size')
# 
# ### cases by enrolment secondary -----------------------------------------------
# idx <- which(cases_per_school$school_level == 'Secondary')
# enrolment_secondary <- as.integer(cases_per_school$school_enrolment[ idx ])
# enrolment_secondary_categorical <- cut(enrolment_secondary, 
# 									   breaks = enrolment_primary_cuts, 
# 									   labels = c('small', 'medium', 'large'))
# case_count_secondary <- as.integer(cases_per_school$cases_per_school[ idx ])
# tapply(case_count_secondary, 
# 	   enrolment_secondary_categorical, 
# 	   sum) %>%
# 	barplot(main = 'infection counts by secondary school size')
# tapply(case_count_secondary / enrolment_secondary, 
# 	   enrolment_secondary_categorical, 
# 	   median) %>%
# 	barplot(main = 'infection rates by secondary school size')
# 
# ### school sizes by school board -----------------------------------------------
# idx <- which(cases_per_school$school_level == 'Secondary')
# enrolment_secondary <- as.integer(cases_per_school$school_enrolment[ idx ])
# enrolment_secondary_categorical <- cut(enrolment_secondary, 
# 									   breaks = enrolment_primary_cuts, 
# 									   labels = c('small', 'medium', 'large'))
# sizes_tbl <- table(cases_per_school$school_board[ idx ],
# 				   enrolment_secondary_categorical)
# # sizes_tbl <- as.data.frame(sizes_tbl)
# large_school_rate <- sizes_tbl[ , 3 ] / rowSums(sizes_tbl)
# 
# ### cases rates by school board ------------------------------------------------
# infection_rates <- cases_per_school$cases_per_school / cases_per_school$school_enrolment
# tapply(infection_rates, 
# 	   cases_per_school$school_board, 
# 	   median) %>% 
# 	sort
```

# Additional Plots

## Policy Events

```{r echo=FALSE, warning=FALSE, message=FALSE}
print(policy_tracing)
```

## Cumulative school-related cases

```{r plot1, echo=FALSE, warning=FALSE, message=FALSE}

# PLOTS ------------------------------------------------------------------------
## school_related_cases_details_plot -------------------------------------------
df <- covid19_schools_summary
idx <- which((df$collected_date >= data_start_date) & (df$collected_date <= data_end_date))
df <- df[ idx, ]
fig <- plot_ly(df, x = ~collected_date, y = ~cumulative_school_related_cases, name = 'Cumulative school-related cases', type = 'scatter', mode = 'lines+markers')
fig <- fig %>% add_trace(y = ~cumulative_school_related_student_cases, name = 'Cumulative school-related student cases', mode = 'lines+markers') 
fig <- fig %>% add_trace(y = ~cumulative_school_related_staff_cases, name = 'Cumulative school-related staff cases', mode = 'lines+markers') 
fig <- fig %>% add_trace(y = ~cumulative_school_related_unidentified_cases, name = 'Cumulative school-related unidentified cases', mode = 'lines+markers')
fig <- fig %>% layout(title = 'Cumulative school-related cases',
					  xaxis = list(title = 'Collected date'),
					  yaxis = list (title = 'Cumulative cases'), 
					  hovermode = 'x unified',
					  legend = list(x = 0.01, y = 0.99, bgcolor = 'rgba(0,0,0,0)'))
event_lines <- lapply(policy_tracing$Date, function(x) {
	list(
		type = 'line', 
		y0 = 0, 
		y1 = 1,
		yref = 'paper', # i.e. y as a proportion of visible region
		x0 = x, 
		x1 = x, 
		line = list(width = 0.25, dash = 'dash', color = 'green')
	)
})
fig <- fig %>% layout(shapes = event_lines)
# fig <- fig %>% add_annotations(x = policy_tracing$Date,
# 							   y = 2500,
# 							   text = format(policy_tracing$Date, '%m-%d'),
# 							   xref = 'x',
# 							   yref = 'y',
# 							   showarrow = TRUE,
# 							   arrowhead = 4,
# 							   arrowsize = .5,
# 							   ax = 20,
# 							   ay = -40)
fig
```

## New school-related cases

```{r plot2, echo=FALSE, warning=FALSE, message=FALSE}
## school_related_new_cases_details_plot ---------------------------------------
df <- covid19_schools_summary
idx <- which((df$collected_date >= data_start_date) & (df$collected_date <= data_end_date))
df <- df[ idx, ]
fig <- plot_ly(df, x = ~collected_date, y = ~new_total_school_related_cases, name = 'New total school-related cases', type = 'scatter', mode = 'lines+markers')
fig <- fig %>% add_trace(y = ~new_school_related_student_cases, name = 'New school-related student cases', mode = 'lines+markers')
fig <- fig %>% add_trace(y = ~new_school_related_staff_cases, name = 'New school-related staff cases', mode = 'lines+markers')
fig <- fig %>% add_trace(y = ~new_school_related_unidentified_cases, name = 'New school-related unidentified cases', mode = 'lines+markers')
fig <- fig %>% layout(title = 'New school-related cases',
					  xaxis = list(title = 'Collected date'),
					  yaxis = list (title = 'New cases'), 
					  hovermode = 'x unified',
					  legend = list(x = 0.01, y = 0.99, bgcolor = 'rgba(0,0,0,0)'))
event_lines <- lapply(policy_tracing$Date, function(x) {
	list(
		type = 'line', 
		y0 = 0, 
		y1 = 1,
		yref = 'paper', # i.e. y as a proportion of visible region
		x0 = x, 
		x1 = x, 
		line = list(width = 0.25, dash = 'dash', color = 'green')
	)
})
fig <- fig %>% layout(shapes = event_lines)
fig
```

## Schools with cases

```{r plot3, echo=FALSE, warning=FALSE, message=FALSE}
## schools_with_cases_plot -----------------------------------------------------
df <- covid19_schools_summary[ , c('collected_date', 'current_schools_with_cases') ]
idx <- which((df$collected_date >= data_start_date) & (df$collected_date <= data_end_date))
df <- df[ idx, ]
fig <- plot_ly(df, x = ~collected_date, y = ~current_schools_with_cases, name = 'Current schools with cases', type = 'scatter', mode = 'lines+markers')
fig <- fig %>% layout(title = 'Schools with cases',
					  xaxis = list(title = 'Collected date'),
					  yaxis = list (title = 'Schools'), 
					  hovermode = 'x unified',
					  legend = list(x = 0.01, y = 0.99, bgcolor = 'rgba(0,0,0,0)'))
event_lines <- lapply(policy_tracing$Date, function(x) {
	list(
		type = 'line', 
		y0 = 0, 
		y1 = 1,
		yref = 'paper', # i.e. y as a proportion of visible region
		x0 = x, 
		x1 = x, 
		line = list(width = 0.25, dash = 'dash', color = 'green')
	)
})
fig <- fig %>% layout(shapes = event_lines)
fig
```

## Active school cases by municipality (top 10)

```{r plot4, echo=FALSE, warning=FALSE, message=FALSE}
## active_cases_by_municipality_plot -------------------------------------------
active_cases_by_municipality <- tapply(covid19_schools_active$municipality,
									   list(covid19_schools_active$collected_date,
									   	 covid19_schools_active$municipality),
									   length)
active_cases_by_municipality <- na.locf(active_cases_by_municipality)
colidx <- order(active_cases_by_municipality[ nrow(active_cases_by_municipality), ], decreasing = TRUE)
active_cases_by_municipality <- active_cases_by_municipality[ , colidx ] 
active_cases_by_municipality <- data.frame(active_cases_by_municipality)
active_cases_by_municipality <- data.frame(collected_date = as.Date(rownames(active_cases_by_municipality)), active_cases_by_municipality)
rownames(active_cases_by_municipality) <- NULL
df <- active_cases_by_municipality[ , 1:11 ]
idx <- which((df$collected_date >= data_start_date) & (df$collected_date <= data_end_date))
df <- df[ idx, ]
nms <- colnames(df)[ -1 ]
nms2 <- colnames(df)[ -1 ] %>% str_replace_all(., '\\.', ' ') %>% str_replace_all(., '\\s+', ' ') %>% str_trim
code_str <- sprintf('
                            fig <- plot_ly(df, x = ~collected_date, y = ~%s, name = \'%s\', type = \'scatter\', mode = \'lines+markers\')
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig
                            ', 
					nms[ 1 ], nms2[ 1 ],
					nms[ 2 ], nms2[ 2 ],
					nms[ 3 ], nms2[ 3 ],
					nms[ 4 ], nms2[ 4 ],
					nms[ 5 ], nms2[ 5 ],
					nms[ 6 ], nms2[ 6 ],
					nms[ 7 ], nms2[ 7 ],
					nms[ 8 ], nms2[ 8 ],
					nms[ 9 ], nms2[ 9 ],
					nms[ 10 ], nms2[ 10 ])
fig <- parse(text = code_str) %>% eval
fig <- fig %>% layout(title = 'Active school cases by municipality (top 10)',
					  xaxis = list(title = 'Collected date'),
					  yaxis = list (title = 'Active cases'), 
					  hovermode = 'x unified',
					  legend = list(x = 0.01, y = 0.99, bgcolor = 'rgba(0,0,0,0)'))
event_lines <- lapply(policy_tracing$Date, function(x) {
	list(
		type = 'line', 
		y0 = 0, 
		y1 = 1,
		yref = 'paper', # i.e. y as a proportion of visible region
		x0 = x, 
		x1 = x, 
		line = list(width = 0.25, dash = 'dash', color = 'green')
	)
})
fig <- fig %>% layout(shapes = event_lines)
fig
```

## Active school cases by school board (top 10)

```{r plot5, echo=FALSE, warning=FALSE, message=FALSE}
## active_cases_by_board_plot --------------------------------------------------
active_cases_by_board <- tapply(covid19_schools_active$school_board,
								list(covid19_schools_active$collected_date,
									 covid19_schools_active$school_board),
								length)
active_cases_by_board <- na.locf(active_cases_by_board)
colidx <- order(active_cases_by_board[ nrow(active_cases_by_board), ], decreasing = TRUE)
active_cases_by_board <- active_cases_by_board[ , colidx ] 
active_cases_by_board <- data.frame(active_cases_by_board)
active_cases_by_board <- data.frame(collected_date = as.Date(rownames(active_cases_by_board)), active_cases_by_board)
rownames(active_cases_by_board) <- NULL
df <- active_cases_by_board[ , 1:11 ]
idx <- which((df$collected_date >= data_start_date) & (df$collected_date <= data_end_date))
df <- df[ idx, ]
nms <- colnames(df)[ -1 ]
nms2 <- colnames(df)[ -1 ] %>% str_replace_all(., '\\.', ' ') %>% str_replace_all(., '\\s+', ' ') %>% str_trim
code_str <- sprintf('
                            fig <- plot_ly(df, x = ~collected_date, y = ~%s, name = \'%s\', type = \'scatter\', mode = \'lines+markers\')
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig <- fig %%>%% add_trace(y = ~%s, name = \'%s\', mode = \'lines+markers\') 
                            fig
                            ', 
					nms[ 1 ], nms2[ 1 ],
					nms[ 2 ], nms2[ 2 ],
					nms[ 3 ], nms2[ 3 ],
					nms[ 4 ], nms2[ 4 ],
					nms[ 5 ], nms2[ 5 ],
					nms[ 6 ], nms2[ 6 ],
					nms[ 7 ], nms2[ 7 ],
					nms[ 8 ], nms2[ 8 ],
					nms[ 9 ], nms2[ 9 ],
					nms[ 10 ], nms2[ 10 ])
fig <- parse(text = code_str) %>% eval
fig <- fig %>% layout(title = 'Active school cases by school board (top 10)',
					  xaxis = list(title = 'Collected date'),
					  yaxis = list (title = 'Active cases'), 
					  hovermode = 'x unified',
					  legend = list(x = 0.01, y = 0.99, bgcolor = 'rgba(0,0,0,0)'))
event_lines <- lapply(policy_tracing$Date, function(x) {
	list(
		type = 'line', 
		y0 = 0, 
		y1 = 1,
		yref = 'paper', # i.e. y as a proportion of visible region
		x0 = x, 
		x1 = x, 
		line = list(width = 0.25, dash = 'dash', color = 'green')
	)
})
fig <- fig %>% layout(shapes = event_lines)
fig

# POLICY CHANGES ---------------------------------------------------------------
# tag plots with policy change events
# visual marker for implementation of asymptomatic testing

```
